---
title: "UKHSA report"
format: html
bibliography: ../bibliography/UKHSA_bibliography.bib
csl: ../bibliography/bmj.csl
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r, echo = FALSE}
#Loading the presentation stuff

source("../R_scripts/source_data_entry.R")
source("../R_scripts/source/swabs.R")
source("../R_scripts/source/primary_care_vis.R")
source("../R_scripts/source/hospitalisation_plots.R")

#For the vaccine plot
source("../R_scripts/source/vacc_plot.R")
```

## Abstract

Influenza is a highly infectious disease caused by a virus of the
*Orthomyxoviridae* family. The virus ( has multiple species, with type A
and type B being predominant in humans. Up to 650,000 deaths yearly can
be directly attributed to seasonal outbreaks of influenza worldwide
(*reference needed*)

Progression of influenza is usually self-limiting. For most, symptoms
last for 2\~7 days. Severe cases might require hospitalisation, or can
progress to death. This is a particular concern for at-risk groups,
which include children \<2 years of age; pregnant women; the elderly
\>65yo; and patients with comorbidities.

The UK's influenza surveillance system is multi-faceted. It includes
sources like weekly % of GP consultations presenting with influenza-like
illnesses (ILI), or death registrations where the leading cause is
influenza. Our topic of interest has been defined with this diversity as
a componenet:

> What is the temporal relationship between different influenza data
> sources (lab-confirmed infections, GP consultations, hospitalisations)
> in the UK, and have these changed in 2022/23 compared to the
> pre-pandemic years 2016-2019?

We hypothesise that there exists a temporal relationship between sources
of surveillance. Statistical tools can then be applied to predict the
timing and scale of a 'later' data source using an 'earlier' source.
Additionally, different age groups can manifest at different time and
scale, and one might be predictive of another. Additionally, we suspect
changes in reporting procedures of respiratory infections post-covid may
cause differences with pre-pandemic data

#### Data Processing

The data used in this project is publicly available on various
governmental websites. Primary care data is published weekly by the RCGP
Research and Surveillance Centre in the communicable and respiratory
disease reports. Laboratory confirmed cases are reported weekly by
UKHSA. Secondary care data is extracted from the UKHSA Severe Acute
Respiratory Infection Watch (SARI Watch) system, updated weekly during
the influenza season (week 40 - week 20 next year). Mortality data is
reported weekly by the ONS. All cleaning and processing are done using
the R language.

#### Analysis

At first glimpse, the peak for each data source does not differ much
from each other in a given year.

Cross-correlation between two data sources can be measured using
Spearman's $ρ$, a non-parametric test of correlation. To ascertain the
pairwise time lag (if any), Spearman's test is run repeatedly with a
gradually increasing lag. The lag value associated with the best $ρ$ is
reported as the lag between given sources for a given year.

*--page break--*

::: {style="page-break-after: always"}
:::

## Introduction

Influenza is a highly infectious disease caused by a virus of the
*Orthomyxoviridae* family. The virus has multiple species, with type A
and type B being predominant in humans. The World Health Organisation
(WHO) estimates that between 290,000-650,000 deaths occur due to the
respiratory virus. This does not include deaths due to secondary
infections caused by influenza [@WHO2022]. Like most viruses, influenza
is highly mutable. It can undergo small mutations (antigenic drift), or
more drastic changes in sub-type (antigenic shift) [@CDC2022]. This can
lead to uncontrolled extensive spreads resulting in an epidemic, such as
the 2009 H1N1 pandemic [@cope2018].

Progression of influenza is often self-limiting. For most, symptoms last
for 2\~7 days. Severe cases might require hospitalisation, or can
progress to death. This is a particular concern for at-risk groups,
which include children \<2 years of age; pregnant women; the elderly
\>65yo; and patients with comorbidities. [@cope2018, @zarocostas2009].

After an incubation period, usually 1\~3 days, the progression of the
disease is self-limiting and symptoms might last 2\~7 days \[cite::green
book\]. In the UK, seasonal influenza circulation is a burden to the NHS
by increasing demand of hospital beds and GP services [@cope2018].
Analysis of previous influenza seasons can help with the planning of
service provision for the future. For example, it would help the
provision of influenza vaccination programmes and temporary bed
expansions to be concordant with the peak of subsequent influenza
seasons, as we'll expand on later.

![**Figure 1:** An infographic summarising the key points addressed in
this report.](images/fig-intro.png){width="92%"}

### Who is the UKHSA and why are they interested

The UK Health Security Agency (UKHSA) was created as a successor to
Public Health England (PHE) in April 2021. Set up during the height of
the COVID-19 pandemic, the UKHSA's main priority is to tackle national
health threats. Their remit includes tackling the influenza epidemic and
managing the nation's vaccine programme.

As mentioned previously, the study of of influenza data (especially its
temporal nature) can help the UKHSA plan for the height of the influenza
season. The data can help forecast the increase in demand for hospital
capacity; and plan the vaccination programme roll out. Vaccination is
greatly beneficial as it prevents between 180,000 and 626,000 cases of
influenza per year in England. Additionally, it prevents between 5,678
and 8,800 premature deaths due to influenza [@franklin2018]. The cost
per death averted by the vaccination programme is estimated to be
£50,610 [@franklin2018].

In the UK, influenza data is reported through three main sources:
primary care data; secondary care data; and laboratory data. Having
three separate data sources is valuable to explore temporal
manifestation of the epidemic between them. This could help forecast a
peak in hospitalisations for example. With the laboratory data, we can
explore the relationship between the strain of influenza and the
hospitalisation rate. Understanding the relationship between age and the
flu epidemic could help tailor the response. For example, if we forecast
an influenza epidemic to disproportionately affect a particular at risk
group, we can tailor the vaccination programme to prioritise those
groups.

### Aims

Our research question, set by the UKHSA, aims to address the issues
outlined above.

*Question:* What is the temporal relationship between different
influenza data sources (lab-confirmed infections, GP consultations,
hospitalisations) in the UK, and have these changed in 2022/23 compared
to the pre-pandemic years 2017-2020?

*Hypothesis:* We suspect that there is a temporal relationship between
one source of surveillance and another. Secondly, we suspect there is
also a relationship between the scale of the peaks between sources of
influenza data. Additionally, we believe there could be changes in
reporting patterns for Influenza-Like-Illnesses (ILI) post COVID.

*Secondary research questions addressed* :

• Does age influence any of the conclusions?

As children have more interactions, we would expect both a larger and
earlier peak as compared with that of the older population.

• What is the impact of different influenza subtypes on
hospitalisations?

We would expect the hospitalisation rate to change depending on which
influenza strain is dominant during the season.

## Methods

### Data sources and extraction

Influenza case data has been extracted from various sources. Firstly,
primary care data was reported by the Royal College of General
Practitioners (RCGP) Research and Surveillance Centre communicable and
respiratory disease reports. Cases of ILI that present in participating
GP practices are collected and reported for England and Wales [@RCGP].

Secondary care data was extracted from the UKHSA Severe Acute
Respiratory Infection (SARI) Watch system from 2022 onwards and from the
UK Severe Influenza Surveillance Schemes (USISS) prior to that.
Lab-confirmed data has been extracted from PHE/UKHSA reports on
respiratory infections.

![**Figure 2:** The data being used comes from various places and has
been sourced through difference
resources.](images/fig-data-source.png){width="100%"}

Mortality data was extracted from the UK Office for National Statistics
(ONS) weekly bulletin of death registrations. Data extracted was defined
by deaths due to or involving influenza as coded by the J09-J18 ICD-10
codes.

Vaccine uptake data is reported monthly by PHE/UKHSA on
[@govukreference]. This data is reported as coverage percentage,
stratified by at risk groups, for the entire season.

### Data Processing

### Statistical Analysis

#### Cross-correlation {#cross-correlation}

To study the temporal relationship between peaks in the influenza data,
a cross-correlation method was chosen. Cross-correlation can be used to
see if two peaks align. The method lags one signal relative to another
and calculates its correlation value. If there is a lag between peaks,
the cross-correlation will result in a higher correlation value for a
peak that has been shifted than for the original peak. This has been
used in addition to rudimentary visual analysis.

To perform this, Spearman's non-parametric test of correlation (denoted
$ρ$) is used. Spearman's $ρ$ uses the ranks of case rate per week to
replace the actual values within each variable. It runs as long as the
variables are equal in length. As seen in Qin *et al.* 2019 [@Qin2019]
The measurement of cross-correlation between pairs of variables should
be as follows. One variable, denoted $x_{t}$, is pinned to the present.
A lag of positive integer $k$ is applied to the other variable $y$ to
generate $y_{t},\; y_{t+1},\;\ldots y_{t+k}$ . For a given influenza
season, Spearman's test is run repeatedly between $x$ and each version
of $y$. The lag that gives the best correlation, denoted
$\arg\underset{k}\max \,\rho(x,y_{t+k})$, is reported for each year for
each pair of variables.

#### Moving Epidemic Method

The Moving Epidemic Method (MEM) is widely used by disease control
agencies (including PHE/UKHSA) to model epidemics and generate threshold
severity levels from historical data. The epidemic period in a season is
the minimum number of consecutive weeks with the maximum accumulated
rates percentage (MAP). The season can then be split into pre-epidemic,
epidemic, and post-epidemic periods [@Vega2013]. Within each, a given
number of data points will be sampled. The number of sample remains at
30 regardless of availability. Assuming these fit into a normal
distribution, their geometric mean will have confidence intervals. The
upper bounds of $\text{CI}_{50},\; \text{CI}_{90},\; \text{CI}_{95}$ act
as the threshold values. Along with the cutoff for entering the epidemic
period, these four values give us five bands: not-epidemic, baseline,
and low, moderate, high, very high threat to the public. In this report,
the plots for GP and hospitalisation use the values from UKHSA
publications for their respective seasons.

#### Reproduction Number $R_0$

Reproduction numbers can be used to calculate rate of disease spread. It
represents the number of new infections one positive case can reproduce.
If $R_0$ is greater than 1, the spread is increasing, while if it is
below 1, the spread is slowing. $R_0$ is typically used at the beginning
of the spread of a new virus to a completely susceptible population. For
a population that contains some immunity, such as for an influenza
virus, an effective reproduction number can be calculated. This can be
inferred from the observed exponential growth rate seen at the beginning
of an influenza epidemic. The observed reproduction number $R$ can be
found through the linear equation $R = 1 + \dfrac{r}{b}$ , where $r$ is
the rate of exponential growth and $b$ is the rate of leaving the
infectious stage [@lipsitch2003]. By approximating the beginning of the
growth curves as $N=ke^{\frac{rt}{b}}$ (where $k$ is a constant), the
value $\dfrac {r}{b}$ can be found from the gradient when plotting
$\log (N)$ against $t$.

In the analysis, log graphs were plotted and a linear model was fitted
to the points at the beginning of the epidemic. Using the gradient, the
$R$ was calculated for each data source and season.

## Results

To get an initial understanding of the trends case rate per 100,000
people was plotted against epidemic week. Lines representing each season
were overlaid on the same plot, MEM thresholds values were visualised
where appropriate.

### Source

::: grid
::: g-col-6
```{r echo= FALSE, warning = FALSE}
gp_cases

```
:::

::: g-col-6
```{r echo= FALSE, warning = FALSE,}
hospital_plot

```
:::
:::

**Figure 3 and 4**: Figure 3 shows RCGP RSC reports of ILI in
participating GP practices per 100,000 people across four influenza
seasons from 2017-2018 to 2022-23 with 2020-2021 and 2021-2022 removed.
Figure 4 show influenza cases reported by SARI watch and USISS Sentinel
data for hospitalisations of patients with influenza per 100,000 people.

```{r echo= FALSE, warning = FALSE}
combined_plot
```

**Figure 5 and 6**: Lab-confirmed cases of influenza, stratified by
influenza strain (A and B). Data reported from PHE/UKHSA lab confirmed
cases of influenza.

In figures 3 and 4, we represent the MEM thresholds against trends of
primary and secondary care cases.

We expected to see that over different seasons, peaks appear at
different weeks. This is due to the complex dynamics of influenza spread
which vary by season. The magnitude of the peaks case rates however
indicates differing levels of severity in influenza between seasons.

This visualisation shows similarity in temporal characteristics between
the 2017-18/2018-19 influenza seasons; and between the 2019-20/2022-23
influenza seasons. In primary care data, the peaks for the 2017-18 and
2018-19 influenza seasons occur at week 4 (with a subsequent one in
2018-19 at week 7). We see a similar split in secondary care data. Case
rate in 2017-18 peaked at week 3, with a smaller one across weeks 8-9.
For 2018-19 the case rate peaks at week 7. In lab confirmed cases data
for influenza A, both 2017-18 and 2018-19 influenza seasons peak at week
7.

In primary care data, the peaks for the 2019-20 and 2022-23 season occur
in the previous year at week 52 and week 50 respectively. In secondary
care data and lab data of influenza A, the peaks for seasons 2019-20 and
2022-23 occur in the previous year at week 52.

Lab confirmed cases of influenza A (Figure 5a) reinforce the back-shift
we see between the 2017-18/2018-19 seasons and the 2019-20/2022-23
seasons. The peak for the 2017-18 and 2018-19 season occur at week 7,
similar to our findings for secondary care and for reported primary care
case rates in the 2018-19 season. The peak case rate for both 2019-20
and 2022-23 seasons occur at week 52, in line with the secondary care
cases for both seasons, and primary care data for 2022-23.

Lab confirmed cases of influenza B only appear to increase notably for
the 2017-18 outbreak, with the cases for subsequent seasons not
increasing to the same magnitude.

It is difficult to be confident in our post-covid analysis as the only
data available covers 2022-23 season, which is still incomplete at time
of reporting. The relationships observed on the brief analysis done may
not hold should the characteristic of 2022-23 season change.

### Season

```{r echo= FALSE, warning = FALSE}
source("../R_scripts/Season_data/season_201718.R")
source("../R_scripts/Season_data/season_201819.R")
source("../R_scripts/Season_data/season_201920.R")
source("../R_scripts/Season_data/season_202223.R")

```

::: grid
::: g-col-6
```{r echo= FALSE, warning = FALSE}
plot_201718
```
:::

::: g-col-6
```{r echo= FALSE, warning = FALSE}
plot_201819
```
:::
:::

::: grid
::: g-col-6
```{r echo= FALSE, warning = FALSE}
plot_201920
```
:::

::: g-col-6
```{r echo= FALSE, warning = FALSE}
plot_202223
```
:::
:::

**Figure 7:** Influenza rates per 100.000, stratified by different
surveillance sources and presented for separate seasons. Data is
combined from RCGP, Datamart scheme and SARI.

```{r echo=FALSE, warning=FALSE}
source("../R_scripts/Season_data/peak_table.R")
max_table

```

For each season, cases of ILI reported by primary care sources surpass
secondary care and lab confirmed case reports. For 2017-18, the peak
value for primary care cases reported by the RCGP was 54.1 cases per
100,000. This was the greatest extent of the influenza epidemic as
reflected by primary care data. The subsequent 2018-19 and 2019-20
seasons saw more modest peaks of 23.1 and 19.4 cases per 100,000
respectively. For the post-COVID influenza season, of data included at
time of reporting, the current peak of primary care cases remains at
31.3 cases per 100,000. This is larger than the two pre-COVID influenza
season data.

Secondary care and lab confirmed case rates mirror each other closely
across pre-pandemic influenza seasons. The magnitude of impact between
hospitalisation and lab-confirmed cases also appear to be similar in
pre-pandemic data.

However, post-COVID, the peak for secondary care data reached 14.8 per
100,000 cases in week 12, whereas the maximum reported rate for lab
confirmed cases reached 5.43 per 100,000 in the same week of data
available. In pre-pandemic influenza seasons, the peaks for secondary
care cases occured during weeks 15 (8.86 per 100,000 in 2017-18), 19
(6.87 per 100,000 in 2018-19) and 12 (7.50 per 100,000 in 2019-20). Lab
confirmed cases peaked during weeks 17 (6.55 per 100,000 in 2017-18), 19
(6.19 per 100,000 in 2018-19) and 12 (6.64 per 100,000 in 2019-20). Its
too early to determine whether this year's influenza season pattern
amongst secondary care and lab confirmed cases matches historical
patterns .

### Log graphs and $R_0$ value

```{r echo= FALSE, warning = FALSE}
source("../R_scripts/Season_data/log_graphs.R")

```

::: grid
::: g-col-6
```{r echo= FALSE, warning = FALSE}
log_hosp_plot1718

```
:::

::: g-col-6
```{r echo= FALSE, warning = FALSE}

log_hosp_plot1819

```
:::
:::

::: grid
::: g-col-6
```{r echo= FALSE, warning = FALSE}

log_hosp_plot1920

```
:::

::: g-col-6
```{r echo= FALSE, warning = FALSE}
log_hosp_plot2223
```
:::
:::

**Figure 8** Log-transformed influenza rates stratified by different
data sources. Results reported for consecutive years.

**Figure 8:** Log-transformed influenza rates stratified by different
data sources. Results reported for consecutive years.

The reproduction number has been estimated for each of the seasons and
each of the data sources. The case rate per 100,000 was transformed
using the natural logarithm and plotted against epidemiological weeks.
The log plot resulted in a straight line during the initial exponential
growth period and a tail after the peak of infections, as was expected
from theory. The plots were visually examined to find which weeks to fit
a linear model to. Using the model, a line was overlaid on the log plots
and coefficients of gradients for each of the lines were calculated. The
$R$ value was obtained from the formula $R = 1 + \frac{r}{b}$ as
detailed above.

The estimated $R$ for each year and data source is detailed below. As
expected at the beginning of a flu season, the R numbers are all
above 1. It is noted R value derived from the hospitalisation data is
higher than that of primary care data.

```{r echo=FALSE}
# source("../R_scripts/Season_data/log_graphs.R")
effective_r_values
```

### Stratifying by age group

```{r echo = FALSE}

source("../R_scripts/Season_data/age_strat.R")
combined_age_plot
```

**Figure 9** Influenza rates per 100.000, stratified by different age
groups.

Using data provided by the RCGP, age stratified rates were plotted
against each other. The unit of measurement is ILI rate per 100,000,
results are broken down into three age groups: \<15 years, 15-64 years
and 65+ years.

For 2017-18, there is a lag between the peaks of adults over 15 years
and children. The rapid growth period of this season begins at week 50
and peaks at week 3 for all ages except \<15 years which peaks later, at
week 5. **This may be due to the timing of the epidemic overlapping with
UK school holidays, leading to children having less social interaction,
therefore delaying the peak for this group by two weeks.**

For 2018-19 there appears to be a minor peak followed by a larger peak a
few weeks later. This year, the second peak among the 65+ age group lags
behind all other age groups by a week. **This may be due to less
socialising among the older age group.**

The season 2019-20 shows multiple peaks with the expected dip over the
holiday period (between weeks 51 and 1). As with the 2018-19 season, the
major peak for 65+ age group lags two weeks behind the other age groups.
There is also an evident spike at week 12 for this year. **As this data
is recording "Influenza like illness", it can be assumed that this peak
is due to the COVID-19 pandemic.**

The current influenza season also shows a lag in growth among the 65+
group. Full season results cannot be presented at the time of reporting,
due to data still being recorded. From the available data we can infer
that the under 15 years have a lower gradient after week 51 around when
the winter holidays begin.

### Stratified by influenza strain type

```{r echo=FALSE}
source("../R_scripts/strain_stratification.R")
combined_strain_plot
```

**Figure 10** Total weekly recorded influenza cases, stratified the
strain as recorded by the Datamart scheme. Presented for consecutive
seasons.

The 2017-18 season was dominated by type B influenza. The earliest rise
in cases was observed in A (H3), starting around week 46, whilst the
other strains present a delay of approximately one week. Moving forward
in the season, week 49 brings an increase in the growth of number of
cases for type B and A (H3). This was the key significant period for
both A (H3) and B type, which went on to reach their peaks in weeks 2
and 3 respectively. The peak number of cases for A (H3) and type B
reached almost 350 and more than 600 cases respectively. Immediately
after its peak, cases of type B influenza started gradually decreasing,
with occasional spikes in weeks 8 and 11, reaching 400 and approximately
250 cases respectively.

A similar trend was observed in the incidence of A (H3) which decreased
in week 6 to 200 cases and fluctuated between 170-250 for the next six
weeks. A (H1N1) starts to plateau around week 2 of 2018, reaching a
point of approximately 50 cases daily. This threshold was not surpassed
for the remainder of the season. The unknown A type reached its peak of
150 cases in week 6. It later fluctuated between 50-100 cases until week
15 when it gradually started to drop.

The data for 2018-19 season only recorded two types of influenza: A
(unknown) and B (unknown). Incidence of type B reached a maximum of 11
cases in the 45th week and did not pass the threshold of 10 cases
thereafter. Growth in the incidence of A (unknown) started at week 44
and continued, with a steeper rise starting in week 51 when over 300 new
cases were recorded. This strain dominated the season soaring at nearly
800 cases in week 3. After a slight dip during the next week, the
incidence of the A (unknown) peaked during week 7 with almost 1250 new
cases. It started plummeting until week 10 and gradually decreased until
the end of the season.

During the 2019-20, cases of A (H1N1), A (unknown) and B (unknown)
caused only a minor burden in comparison to A (H3). All of the analysed
strains apart from A (H3) fluctuate around 10-30 cases between weeks 43
of 2019 and 9 of 2020. The incidence of A (H3) started to rise during
week 43 of the season and peaked in week 52. After its peak, it dropped
rapidly to eventually level off during week 5 and fluctuated around the
same incidence.

For the 2022-23 season we can observe that both type B and A (H1N1)
represented only a fraction of all incidence, reaching approximately 50
cases each during the 52nd week. The most dominant strain was A
(unknown), which was on the rise since the first week of the season.
During the 50th week, cases started soaring until the incidence of the
type of influenza reached its peak of almost 2300 new cases in week 52
of 2022. A (H3) was the second most dominant strain. It presented a
slower rise, starting in week 46 and peaking at 1250 new cases. Type A
(unknown) also peaked at week 52. Data in the graph only represents
incidence until week 5 of the year 2023 and that is when we can observe
a decline in all of the cases.

## Discussion

### Main piece of supporting evidence

From analysis of the plots for each season stratified by surveillance
source, the peaks for rate of cases line up. The sample size of four
seasons is small, it appears there is no temporal relationship between
different influenza data sources. While we see no visible lag between
peaks in the population as a whole, the data was probed further to try
to discover any other causal inference.

### **Interpretation of the corr**

From the results calculated, it is clear that the estimate is larger for
both hospitalisation data and swab data than for the GP data. The
calculation here is a simplistic approximation for the reproduction
number but does show that the growth of cases is lower for GP
consultations. This is more indicative of the general population as not
everyone who goes to a GP for ILI symptoms will be referred to hospital.
This suggests some evidence towards a temporal relationship between
primary and secondary care data, as knowing when there is a rise in the
cases in GPs infers a subsequent large rise in hospitalisations. The
calculation is not a complex method of estimating the reproduction
number for influenza but does give a operationalised approximation.
Estimating the reproduction number for a virus such as influenza can be
difficult due to uncertainty in the level of immunity among the
population.

### Strain stratified data discussion

Stratification of the results by different strains has brought some
interesting insights into our study. The 2017-18 season was dominated by
type B influenza. This finding is particularly interesting as according
to the WHO \[7\], the northern hemisphere has recorded an increased
incidence of A (H1N1), which in the UK was only a secondary strain.
Another interesting point is that the peak incidence of type B influenza
was delayed in the other countries of the northern hemisphere during the
2017 season. Within the UK, we have noted an earlier peak of A (H1N1)
compared to type B. During the season 2018-19, data was only stratified
by type and not subtype. According to the WHO, this season have been
dominated by A (H1N1) in the northern hemisphere. Compared to the data
from northern Europe, we can see similarities in the shape distribution
of lab-confirmed cases. Other strains of influenza were not noticeably
present. In the 2019-20 season, we observed dominance of A (H1N1).
Contrast this with the rest of northern Europe, where the most prevalent
strains were type A and B of unknown lineages. The peak in the UK for A
(H1N1) was delayed in comparison to the cumulative peak for northern
Europe. It should be noted that the UK peak lined up with the peak for
the A (H3) and type B for northern Europe. Incidence of A (H3) and type
B in the UK was delayed compared to A (H1N1). However, the general
incidence of A (H3) and B was minimal throughout the whole season.
Season 2022-23 started early, with the incidence of A (unknown) leading
the rising trend. Results correspond with data for northern Europe,
sharing the shape of the distribution and dominating strain. It is noted
that this is the first season in our observed timeframe where UK data is
consistent with data for northern Europe. In comparison with data for
the countries of the whole northern hemisphere, the incidence in the UK
seems to plummet with a slight delay.

### Comparison and contrast to previous studies

Few studies have been done on this topic, and none on UK data. However,
there are several sources which try to understand the relationship of
influenza before and after the COVID-19 pandemic. One of the most
interesting comparisons would be with the work by [@groves2022] studying the
effect of the COVID-19 pandemic on ILI-related hospitalization and
mortality in children [@groves2022]. The main findings conclude there was a
statistically significant drop in the number of cases in pediatric
influenza hospitalisations post pandemic. The report highlights the lack
of ILI-related data for the COVID-19 pandemic period, in line with our
findings. A similar phenomenon is seen in Italy, China and the USA
[@young2020]. It is suspected that during that period, ILI was under-reported
due to the main focus being COVID testing. This hypothesis however does
not explain observations in China where incidence was low, even though
type A and B influenza tests were conducted for patients with suspicion
of COVID-19 [@huang2020]. Another possible explanation is the change in contact
patterns which caused immunity perturbations that happened throughout
2019-20.

Seasonality is a result of a complex interplay between previously
mentioned contact patterns, host immunity and survival of the virus
[@tamerius2021]. In terms of the post-pandemic period, sources indicate that the
coming season carries multiple unpredictable factors such as the lasting
effect of the COVID-19 pandemic, decreasing population immunity and a
higher genetic homogeneity amongst influenza lineages [@dhan2022].
Inconsistent COVID vaccination patterns i.e. varied coverage depending
on a country, group age and time window, disrupted the roll-out of other
vaccines [@maltzou2022]. This shows some similarities with our findings, as seen
in the appendix (Fig. Vacc). Uptake rate in most risk groups dropped in
2020 compared to 2019. Pregnant women were the only risk group with
similar uptake rate. These disruptive patterns were observed in South
Africa when the 2021-22 season peaked unusually late during the winter
[@whoupdate]. However, regular seasonality was restored in the following
season, both dominated by A (H1N1).

### Strengths

One of the strong points of our report was the availability of multiple
data sources. As opposed to similar literature, we have gathered data
covering separate sources (lab-confirmed tests, primary and secondary
care) in order to establish temporal relations between them. Our report
contributes to currently available knowledge by presenting insight on
the differences in reporting patterns between surveillance sources.
Methods, such as $R_0$, allowed us to compare the rates of epidemic
growth between different data sources. Along with the stratification by
age groups, we get a better picture which can be used as decision
rationale for better allocation of public resources.

### Limitations

Inevitably our research is limited by only using publicly available
data. Our data did not have the quantity and quality needed for more
in-depth statistical analysis. For example, Pearson Correlation test
would require the data to be normally distributed, and Granger Causality
needs prerequisite tests to prove stationarity. We resorted to use
Spearman Correlation with manual lag ([\#](#Cross-correlation)), which
would result in `NA`s in the first few or last few observations. The
test has to be run on two series of equal length, meaning that a
considerable amount is trimmed off to match them. Lack of more granular
data is another drawback. While it does not impact the calculation of
$R_0$ values, having measurements only on a weekly scale made it
impossible to compare peaks that might come a few days after one
another. From what is available, we concluded counter-intuitively that
the peak younger group did not lead that of the elderly group, reasons
unclear.

Throughout this project, we have not attempted, nor it is part of the
agenda to forecast flu. While there exists machine learning methods
(e.g. ARIMA) to make reasonable forecasts, in-depth understanding of
infection dynamics is needed to build a suitable *A Priori* model.
Considering all the above, we can state that there is room for
improvement with future research on this topic.

### Hypothesis and purpose of study

Our hypothesis was that we expected there to be a temporal difference
between peaks from different data sources. We suspected that
hospitalisations may lag behind GP cases. When stratifying results by
season according to the source, there is no clear change in the timing
of the peaks with the exception of 2022-23. Within the same season, the
maximum rate of cases occur in the same week. We hypothesised a lag
leading to hospitalisation on the basis that influenza takes time to
develop severe symptoms. It is known that the incubation period of
influenza is between 0.6 and 1.4 days and most people will recover
within a week [@Lancet2022]. Due to this, it is not surprising that
there is no obvious lag between data sources as the time frame is not
small enough to capture any differences between peaks. Any delay in
reporting cases from any of the sources would make it more difficult to
distinguish any differences in peaks. Therefore, we cannot say there is
a temporal relationship present in the data. GP consultations for ILI
are not appropriate for forecasting subsequent hospitalisation cases.

With regards to the secondary research questions, the data gives more
insight. When GP consultations are stratified by age group, there is a
clear lag between the 65+ group and the remaining age groups in all
seasons except 2017-18. The one season where \<15 group lags behind the
other age groups can be explained as the peak of infections happening
after the holiday period when schools are closed. When schools are
closed, younger people have fewer social contacts and are therefore less
likely to spread influenza. A delayed peak occurring after the return to
school, as shown in the plots, is to be expected. Referring back to the
2018-19, 2019-20 and 2022-23 seasons, the lag of the age group 65+ is
also to be expected. For a similar reasoning, 65+ age group will tend to
have fewer social interactions than children, so will have a slower
spread of influenza. Therefore, knowing that cases are rising among \<15
years can help predict a peak in the 65+ age group, who also would tend
to be more severely affected by influenza.

#### Future work

It has been reported in the past that different strains of influenza may
have a different distribution between different age groups [@Age2015].
It could be useful in the future to have cases reported by both strain
and age group so that any differences can be studied. This may also
contribute to a temporal difference and could help prepare for rise in
cases as some strains have a higher hospitalisation rate than others.

As COVID-19 becomes endemic within the population, it will be important
for the UKHSA to know the temporal relationship of both diseases. COVID
concominant with influenza has the potential to increase secondary care
burden. Therefore, studying a temporal relationship of influenza and
COVID-19 for each data source could help epidemic preparedness.

Due to our data sources being collated by the UKHSA, the data used
within this report is predominantly from England. Influenza circulates
around the world and is tracked by many international bodies. While
acknowledging the different structure of the UK public health
infrastructure, it may also be worth studying temporal differences
between influenza data sources in other countries.

## Conclusion

Influenza is a very difficult disease to predict due to reasons
including the different strains, antigenic shift and the unknown level
of immunity in a given season. While data for previous seasons can be
studied, the complex nature and season to season variability mean it is
difficult to give a firm prediction of what could happen in the future.
Although we see no noticeable lag in peaks between any of the data
sources, there is a lag in primary care cases among different age
groups. There is also a noticeable difference in the rate of spread
reported between primary and secondary care cases.

## Appendix A - Team Statement

-   Data extraction and cleaning: GB, AC, SJ, HL

-   Scripting: GB, AC, SJ, HL

-   Statistical Analysis: GB, HL

-   Visualisation: GB, AC, SJ, HL

-   Writing-Up: GB, AC, SJ, HL


I hereby declare and confirm with my signature that this report, along
with the code and graphics in this project, is the combined effort of
our group. Everyone has made significant contributions in their own
ways, based on their expertise in one speciality or another. I
acknowledge that this report is submitted as coursework to be assessed,
and the team members are equally responsible for the success or failure
of any one part. This report is my own work, except where the use of
other people's works are cited.

Team Member: GB, AC, SJ, HL

## **Appendix B - Variables and Data Sources**

All data sources and scripts can be found in our shared github :
<https://github.com/gabrielbattcock/data_challenge>.

## Appendix C - Summary of Missing Data

## References
