---
title: "Mid Module Presentation"
format: 
  revealjs:
    theme: beige
    incremental: true
    menu: 
      side: left
      width: third
editor: visual
bibliography: UKHSA_bibliography.bib
csl: bmj.csl
---

# Introduction

```{r}
#Loading the presentation stuff
source("source_data_entry.R")
source("swabs.R")
#For the vaccine plot
source("vacc_plot.R")

```

## Background {.scrollable}

-   Highly infectious and rapidly evolving respiratory virus circulating seasonally around the world [@cope2018]
-   Antigenic shifts may result in uncontrolled extensive spread causing an epidemic (e.g. 1918-2009) [@cope2018]
-   Progression may lead to deaths in at risk groups (children \<2 y.o., pregnant women, people \>65 y.o. and people with underlying medical conditions) [@zarocostas2009]
-   Seasonal influenza circulation puts a burden on NHS e.g. emergency room and GP services [@cope2018]
-   Vaccination averts between 180,000 and 626,000 cases of influenza per year in England [@franklin2018]
-   Flu vaccination helps avert between 5,678 and 8,800 premature deaths per year [@franklin2018]
-   The NHS flu vaccination programme costs £50,610 per death averted [@franklin2018]
-   Understanding the circulation patterns can help UKHSA:
    -   effectively allocate resources
    -   forecast future season dynamics
    -   to aid early detection of anomalous seasons

## Our Question

What is the temporal relationship between different influenza data sources (lab-confirmed infections, GP consultations, hospitalisations) in the UK, and have these changed in 2022/3 compared to the pre-pandemic years 2016-2019?

## Data Sources

::: incremental
-   **Primary care**: Royal College of GP appointments for Influence Like Illnesses
-   **Secondary care**: Hospitalisation data as reported by th UKHSA Severe Acute Respiratory Infection (SARI) Watch system and the UK Severe Influenza Surveillance Schemes (USISS) prior to that.
-   **Laboratory confirmed**: Reports of respiratory infections made to PHE from PHE and NHS laboratories in England and Wales
:::

# Methods


## Data aquisition

-   Various methods of converting data to a usable form have been used:
    -   UKHSA provided OpenDocument spreadsheets (.ODS) for all flu data (RCGP, USISS and SARI) from 2018-23
    -   PDF scraping (use of R script provided to us) - swabs
    -   2017-18 data for hospitalisation and GP consultations was manually entered from weekly published reports.

## Analysis

-   Time series plots
-   $R_0$ value for early spread
-   Calculate delay of peak/initial spread of flu between different data sources


# Results

## Exploring our main data sources

::: panel-tabset
### Primary care

```{r}
source("primary_care_vis.R")
gp_cases
```

### Secondary Care

```{r}
source("hospitalisation_plots.R")
hospital_plot
```

### Lab confirmed cases

```{r}
source("swabs.R")
combined_plot
```
:::

## Lab confirmed influenza cases

::: panel-tabset
### Influenza A subtype

```{r}
plotA
```

### Influenza B subtype

```{r}

plotB
```
:::

## Vaccine data

```{r}
vacc_plot
```

## Mortality Data

## Per flu seasons

::: panel-tabset
### 2017-18

```{r}
source("season_201718.R")
plot_201718
```

### 2018-19

```{r}
source("season_201819.R")
plot_201819
```

### 2019-20

```{r}
source("season_201920.R")
plot_201920
```

### 2022-23

```{r}
source("season_202223.R")
plot_202223
```
:::

## Next steps

-   Calculating R0 values
-   Calculating time of peaks
-   Calculating delay between onset of season and peaks
-   Stratifying results by age

## References
